For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-01 paper/paper.tex
stuff

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by Thematic Excellence Programme, Industry and Digitization Subprogramme (NRDI Office, 2019) and by the European Union, co-financed by the European Social Fund (EFOP-3.6.2-16-2017-00013, Thematic Fundamental Research Collaborations Grounding Innovation in Informatics and Infocommunication).}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}

@@

First we define the term signature algebra using which we obtain the
constructors $\con$ for any IIT signature. Then we will define another
signature algebra which provides the eliminator. Before doing these,
we illustrate the idea on natural numbers.
\begin{example}
  For natural numbers, we will define the constructors $\con$ as the
  following natural number algebra
  $(\mathsf{Nat},\mathsf{zero},\mathsf{suc})$. We write variable names
  instead of de Bruijn indices for readability.
  \begin{alignat*}{5}
    & \mathsf{Nat} && := \Tm^\I\,(\cdot\rhd N:\U\rhd z:\El\,N\rhd s:N\Ra\El\,N)\,(\El\,N) \\
    & \mathsf{zero} && := z \\
    & \mathsf{suc} && := \lambda t. (s\oldapp t)
  \end{alignat*}
  Natural numbers are simply $\I$-terms of type $\El\,N$ in the
  context representing the signature for natural numbers. In this
  context, the only way to define a term of type $\El\,N$ is to use
  $z$ and $s$, corresponding to the $\mathsf{zero}$ and $\mathsf{suc}$
  constructors.

@@

  We define ${\A M\C}_\Omega$ by listing all its components $\Con$,
  $\Ty$, $\Sub$, and so on, one per row. Each component has three
  parts denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are
  just reusing the corresponding parts from $\A$ and $M$,
  respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
  defined as the obvious projections. We omit the equality components,
  as they come from UIP or are trivial. We also omit the components
  for terms and substitutions as their $^\C$ parts are uninformative
  equational reasonings.
  \begin{alignat*}{10}
    & \Gamma^\A:\Set && \hspace{0.4em} && \Gamma^M:\Con^M && \hspace{0.4em} && \Gamma^\C: \Sub^M\,\Omega\,\Gamma^M\ra \Gamma^\A \\
    & A^\A:\Gamma^\A\ra\Set && && A^M:\Ty^M\,\Gamma^M && && A^\C:(\nu:\Sub^M\,\Omega\,\Gamma^M)\ra \\
    & && && && && \hspace{1em} \Tm^M\,\Omega\,(A^M[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
    & \sigma^\A:\Gamma^\A\ra\Delta^\A && && \sigma^M:\Sub^M\,\Gamma^M\,\Delta^M && && \sigma^\C:\Delta^\C\,(\sigma^M\circ\nu)=\sigma^A\,(\Gamma^\C\,\nu) \\
    & t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma && && t^M:\Tm^M\,\Gamma^M\,A^M && && t^\C : A^\C\,\nu\,(t^M[\nu])= t^A\,(\Gamma^\C\,\nu) \\
    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^M := A^M[\sigma^M]^M && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^M\circ\nu)\,t \\
    & \cdot^\A := \top && && \cdot^M := \cdot^M && && \cdot^\C\,\nu := \tt \\
    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^M :=  && && (\Gamma\rhd A)^\C\,\nu := \\
    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} \Gamma^M\rhd^M A^M && && \hspace{1em} (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
    & \U^\A\,\gamma := \Set && && \U^M := \U^M && && \U^\C\nu\,a := \Tm^M\,\Omega\,(\El^M\,a) \\
    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^M := \El^M\,a^M && && (\El\,a)^\C\,\nu\,t := (a^\C\,\nu) \# t \\
    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^M := && && (\Pi\,a\,B)^\C\,\nu\,t :=  \\
    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em} \Pi^M\,a^M\,B^M && && \hspace{1em} \lambda\alpha.B^\C\,\big(\nu, (a^\C\,\nu)\#\alpha\big)\,\big(t\oldapp ((a^\C\,\nu) \#\alpha)\big) \\
    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^M := && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha) \\
    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} \Pim^M\,T\,B^M && && 
  \end{alignat*}
\end{definition}

xxx

>>>>

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no. ED18-1-2019-0030 (Application-specific highly reliable IT
solutions), by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no. ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no. BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}

@@

\begin{remark}
  In a model of ETT supporting the theory of ITT signatures, we can
  view elements of $\Con^\I$ as syntactic signatures. We can turn a
  $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$ and an
  $\Omega:\Con^\I$ into a $\Sign$ by
  $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
  \ll\Omega\rr_N)$ where the equality in the second component comes
  from uniqueness of the recursor.
\end{remark}

First we define the term signature algebra using which we obtain the
constructors $\con$ for any IIT signature. Then we will define another
signature algebra which provides the eliminator. Before doing these,
we illustrate the idea on natural numbers.
\begin{example}\label{eg:nat_con}
  For natural numbers, we will define the constructors $\con$ as the
  following natural number algebra
  $(\mathsf{Nat},\mathsf{zero},\mathsf{suc})$. We write variable names
  instead of de Bruijn indices for readability.
  \begin{alignat*}{5}
    & \mathsf{Nat} && := \Tm^\I\,(\cdot\rhd N:\U\rhd z:\El\,N\rhd s:N\Ra\El\,N)\,(\El\,N) \\
    & \mathsf{zero} && := z \\
    & \mathsf{suc} && := \lambda t. (s\oldapp t)
  \end{alignat*}
  Natural numbers are simply $\I$-terms of type $\El\,N$ in the
  context which is the syntactic signature for natural numbers. In
  this context, the only way to define a term of type $\El\,N$ is to
  use $z$ and $s$, corresponding to the $\mathsf{zero}$ and
  $\mathsf{suc}$ constructors.

@@

  We define ${\A M\C}_\Omega$ by listing its components $\Con$, $\Ty$,
  $\Sub$, and so on, one per row. Each component has three parts
  denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are just
  reusing the corresponding components from $\A$ and $M$,
  respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
  defined as the obvious projections. We omit the equality components,
  as they come from UIP or are trivial. We also omit the components
  for terms and substitutions as their $^\C$ parts are uninformative
  equational reasonings.
  \begin{alignat*}{10}
    & \Gamma^\A:\Set && \hspace{0.4em} && \Gamma^M:\Con^M && \hspace{0.4em} && \Gamma^\C: \Sub^M\,\Omega\,\Gamma^M\ra \Gamma^\A \\
    & A^\A:\Gamma^\A\ra\Set && && A^M:\Ty^M\,\Gamma^M && && A^\C:(\nu:\Sub^M\,\Omega\,\Gamma^M)\ra \\
    & && && && && \hspace{1em} \Tm^M\,\Omega\,(A^M[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
    & \sigma^\A:\Gamma^\A\ra\Delta^\A && && \sigma^M:\Sub^M\,\Gamma^M\,\Delta^M && && \sigma^\C:\Delta^\C\,(\sigma^M\circ\nu)=\sigma^A\,(\Gamma^\C\,\nu) \\
    & t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma && && t^M:\Tm^M\,\Gamma^M\,A^M && && t^\C : A^\C\,\nu\,(t^M[\nu])= t^A\,(\Gamma^\C\,\nu) \\
    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^M := A^M[\sigma^M]^M && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^M\circ\nu)\,t \\
    & \cdot^\A := \top && && \cdot^M := \cdot^M && && \cdot^\C\,\nu := \tt \\
    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^M :=  && && (\Gamma\rhd A)^\C\,\nu := \\
    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} \Gamma^M\rhd^M A^M && && \hspace{1em} (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
    & \U^\A\,\gamma := \Set && && \U^M := \U^M && && \U^\C\nu\,a := \Tm^M\,\Omega\,(\El^M\,a) \\
    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^M := \El^M\,a^M && && (\El\,a)^\C\,\nu\,t := (a^\C\,\nu) \# t \\
    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^M := && && (\Pi\,a\,B)^\C\,\nu\,t :=  \\
    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em} \Pi^M\,a^M\,B^M && && \hspace{1em} \lambda\alpha.B^\C\,\big(\nu, (a^\C\,\nu)\#\alpha\big)\,\big(t\oldapp ((a^\C\,\nu) \#\alpha)\big) \\
    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^M := && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha) \\
    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} \Pim^M\,T\,B^M && && 
  \end{alignat*}
\end{definition}
\begin{example}
  Now given a syntactic signature $\Omega:\Con^\I$, we get the
  constructors as an $\Omega$-algebra by
  $(\ll\Omega\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \ll\Omega\rr^\A$. If
  $\Omega$ is the syntactic signature for natural numbers, we get the
  constructors as in \Cref{eg:nat_con}.

  An $a : \Tm\,\Omega\,\U$ is a sort term for a signature $\Omega$. If
  $\Omega$ is the syntactic signature for natural numbers, $a$ can
  only be $N$ ($1$ as a de Bruijn index). If $\Omega$ is the syntactic
  signature for $\Con$--$\Ty$ (\Cref{ex:signatures}), $a$ can be
  $Con$, $Ty\oldapp{empty}$,
  $Ty\oldapp({ext}\oldapp {empty}\oldapp(U\oldapp{empty}))$, and so
  on. In any case, for such an $a$, we obtain
  $(\ll a\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a)
  = \ll a\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, the algebra
  interpretation of a sort term at the constructors is equal to terms
  of that sort.

  A $t : \Tm\,\Omega\,(\El\,a)$ is a term of a sort type $a$
  constructed using the constructors in $\Omega$. For natural numbers,
  such a $t$ can only be $s$ applied iteratively to $z$. For such a
  $t$, we obtain
  $\ll t\rr^\C\,\id^\I : t = \ll
  t\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, a constructor term is
  equal to its algebra interpretation at the constructors. This is
  exactly the equation needed at the end of \Cref{eg:nat_con}.
\end{example}
\begin{remark}
  Given a syntactic signature $\Omega : \Con^\I$, we can use the term
  signature algebra with choices other than $M = \I$. For example, if
  $M = \A$, we usual get the Church encoding by
  $(\ll\Omega\rr_{{\A\A\C}_{\ll\Omega\rr^\A}})^\C\,\id^\A$. In the
  case of natural numbers, this computes (up to curry-uncurry) to
  $(N:\Set)(z:N)(s:N\ra N)\ra N$. Using $M = {\A\M}$ (the signature
  algebra $\A$ extended with the binary version of $\D$ also called
  logical relations, see \cite[Section
  5]{Kaposi:2019:CQI:3302515.3290315}), we obtain the Church encoding
  of Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}.
\end{remark}

\begin{definition}[Eliminator signature algebra $\IE_{\blank}$]
  Given an $\Omega:\Con^\I$,
  $\omega^D : \ll\Omega\rr^\D\,(\ll\Omega\rr^\C\,\id^\I)$, we define a
  the signature algebra $\IE_{\omega^\D}$ as follows. It is equipped
  with a morphism $\blank^\I : \SignMor\,\IE_{\omega^D}\,\I$.

  We define $\IE_{\omega^D}$ by listing its components $\Con$, $\Ty$,
  $\Sub$, and so on, one per row. Each component has two parts denoted
  by $^\I$, and $^\E$. The $^\I$ part just reuses the corresponding
  components of $\I$, thus the morphism $\blank^\I$ is defined as the
  obvious projection. We omit the equality components, as they come
  from UIP or are trivial. We also omit the components for terms and
  substitutions as their $^\E$ parts are uninformative equational
  reasonings.

xxx
\end{definition}
